---
title: "R Zone 4: Chapter 12 and 15"
author: " Brandon Hufstetler, Garrett Alarcon, Jinan Andrews, Anson Cheng, Nick Forrest, and Nestor Herandez"
date: "5 August 2019"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
  hmtl_document: default
geometry: margin = 1in
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyfoot[CO,CE]{Forrest - Page \thepage}
fontsize: 11pt
fontfamily: mathpazo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

# Chapter 12

### READ IN AND PREPARE THE DATA
```{r}
  setwd("~/IMGT680")
  adult <- read.csv(file = "adult.txt", stringsAsFactors=TRUE)
  
  # Collapse categories as in Chapter 11
  # We will work with a small sample of data
  adult <- adult[1:500,]
```

### DETERMINE HOW MANY INDICATOR VARIABLES ARE NEEDED
#### Here we are looking at how many different unique levels(categories) are present within each feature in the data. This shows that there are 2 different levels of income, 2 different levels of gender, 5 different levels of race, 9 different levels of workclass, and 7 different levels of marital status in the adult.txt dataset. 
```{r}
  unique(adult$income)             # One variable for income
  unique(adult$sex)                   # One variable for sex
  unique(adult$race)                  # Four variables for race
  unique(adult$workclass)         # Three variables for workclass
  unique(adult$marital.status)   # Four variables for marital.status
```

### CREATE INDICATOR VARIABLES
#### First, we define the new indicator variables that we want to populate. These will all be binary indicators, so if a specific level is present in the data, the value of the indicator is equal to 1. The levels in this data are mutually exclusive, so only one level from each feature can occur in one observation. Therefore, it is not necessary to make an indicator variable for every level contained in a feature. The level that is left out is the assume baseline, as it is the only other option given the other indicator variables from that feature are 0. 
```{r}
  adult$race_white <- adult$race_black <- adult$race_as.pac.is <-
  adult$race_am.in.esk <- adult$wc_gov <-  adult$wc_self <- adult$wc_priv <-
  adult$ms_marr <- adult$ms_div <- adult$ms_sep <- adult$ms_wid <-
  adult$income_g50K <- adult$sex2 <- c(rep(0, length(adult$income)))
  for (i in 1:length(adult$income)) {
    if(adult$income[i]==">50K.")
      adult$income_g50K[i]<-1
    if(adult$sex[i] == "Male")
      adult$sex2[i] <- 1
    if(adult$race[i] == "White") adult$race_white[i] <- 1
    if(adult$race[i] == "Amer-Indian-Eskimo") adult$race_am.in.esk[i] <- 1
    if(adult$race[i] == "Asian-Pac-Islander") adult$race_as.pac.is[i] <- 1
    if(adult$race[i] == "Black") adult$race_black[i] <- 1
    if(adult$workclass[i] == "Gov") adult$wc_gov[i] <- 1
    if(adult$workclass[i] == "Self") adult$wc_self[i] <- 1
    if(adult$workclass[i] == "Private" ) adult$wc_priv[i] <- 1
    if(adult$marital.status[i] == "Married") adult$ms_marr[i] <- 1
    if(adult$marital.status[i] == "Divorced" ) adult$ms_div[i] <- 1
    if(adult$marital.status[i] == "Separated" ) adult$ms_sep[i] <- 1
    if(adult$marital.status[i] == "Widowed" ) adult$ms_wid[i] <- 1
  }
```  

### MINIMAX TRANSFORM THE CONTINUOUS VARIABLES
#### Standardization is necessesary to use continuous variables as inputs for neural networks. Here we use mix-max normalization to transform the continuous data into values between 0 and 1. Normalization is necessary because otherwise, the model would favor the larger values when determining weights for each node in the model during back propogation. Non-normalized data would lead to inaccurate predictions on new data. 
```{r}
  adult$age_mm <- (adult$age - min(adult$age))/(max(adult$age)-min(adult$age))
  adult$edu.num_mm <- (adult$education.num - min(adult$education.num))/
    (max(adult$education.num)-min(adult$education.num))
  adult$capital.gain_mm <- (adult$capital.gain - min(adult$capital.gain))/
    (max(adult$capital.gain)- min(adult$capital.gain))
  adult$capital.loss_mm <- (adult$capital.loss - min(adult$capital.loss))/
    (max(adult$capital.loss)- min(adult$capital.loss))
  adult$hours.p.w_mm <- (adult$hours.per.week - min(adult$hours.per.week))/
    (max(adult$hours.per.week)-min(adult$hours.per.week))
  newdat <- as.data.frame(adult[,-c(1:15)]) # Get rid of the variables we no longer need
```

### RUN THE NEURAL NET
### This neural net aims to predict whether or not the income of a new adult is greater than $50k. The weights are the parameters being fit to optimize the model. Unless a random seed is set, the weights are going to be different everyt time we run this code. This is because the fitting/training process may start from a different point eveyr time. This particular network has an input layer, a hidden layer, and an output layer, the hidden layer has 8 nodes. R defaults to using a log-linear activation function within each of these nodes. There is only 1 ouput node. 
```{r}
  #install.packages("nnet")
  library(nnet) # Requires package nnet
  net.dat <- nnet(income_g50K ~ ., data = newdat, size = 8)
  table(round(net.dat$fitted.values, 1))   # If fitted values are all the same, rerun nnet
  net.dat$wts # Weights
  hist(net.dat$wts)
```

# Chapter 15
  
### THE CONFUSION MATRIX
  # After using the C5.0 package, the confusion matrix is included in the output of summary()
  # See Chapter 11 for data preparation and code to implement the C5.0 package
#### The confusion matrix is a good way visualize the performance of the model in a table. A confusion matrix compares the predicted results to the actual results. This gives us useful information to calculate metrics such as True Positive Rate, False Positive Rate, Accuracy, Precision, and F-Score. These metrics are ways to determine the success of the algorithm depending on the contexts of the problem.  
### Chapter 11 data prep
```{r}
  setwd("~/IMGT680")
  adult <- read.csv(file = "adult.txt", stringsAsFactors=TRUE)
  
  #Collapse some of the categories by giving them the same factor label
  levels(adult$marital.status);levels(adult$workclass)
  levels(adult$marital.status)[2:4]<-"Married"
  levels(adult$workclass)[c(2,3,8)]<- "Gov"
  levels(adult$workclass)[c(5,6)] <- "Self"
  levels(adult$marital.status); levels(adult$workclass)
  
  #add columns with normalized data
  adult$age.z <- (adult$age - mean(adult$age))/sd(adult$age)
  adult$education.num.z <- (adult$education.num - mean(adult$education.num))/sd(adult$education.num)
  adult$capital.gain.z<- (adult$capital.gain - mean(adult$capital.gain))/sd(adult$capital.gain)
  adult$capital.loss.z <- (adult$capital.loss - mean(adult$capital.loss))/sd(adult$capital.loss)
  adult$hours.per.week.z <- (adult$hours.per.week - mean(adult$hours.per.week))/ sd(adult$hours.per.week)
                                                                    
```

### ADD COSTS TO THE MODEL
```{r}
  #install.packages("C50")
  
  #After data preparation from Chapter 11
  x <- adult[,c(2,6, 9, 10, 16, 17, 18, 19, 20)]
  y <- adult$income
  
  # Without weights:
  c50fit <- C5.0(x, y, control = C5.0Control(CF=.1))
  summary(c50fit)
```

```{r}
  # With weights:
  costm <- matrix(c(1, 2, 1, 1), byrow = FALSE, 2, 2)
  c50cost <- C5.0(x, y, costs = costm, control = C5.0Control(CF=.1))
  summary(c50cost)
```
 