---
title: "R Zone 1: Chapter 1 and 2"
date: "1 July 2019"
author: " Brandon Hufstetler, Garrett Alarcon, Jinan Andrews, Anson Cheng, Nick Forrest, and Nestor Herandez"
output: 
  # tufte_handout:
    # highlight: tango
  # html_document:
  #   theme: journal
  tufte::tufte_html:
    keep_md: true
    # highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

# Chapter 1

### GETTING STARTED WITH R
### Comments, indents, and semicolons
```{r}
  # Anything prefaced by a pound sign (#) is a comment.
  # Comments are not executed by R. Instead, they explain what the code is doing.
  # Indented code (that is not a comment) will run in R as if it was on one line
  # Code separated by semicolons will run as if the code was on separate lines,
  # with the semicolon marking the line break
```

### Open a dataset and display the data
#### Commentary: The commands below read and output the "Cars" data set as well as specific portions of the data set for the user to see.
```{r,include = FALSE}
  # Replace “C:/…/” with the exact location of the file you want to open
  filepath <- ""
  setwd("~/IMGT680") 
  cars <- read.csv(file = "cars.txt", stringsAsFactors = FALSE)
  cars # To display the whole dataset, type the dataset name
```

```{r}
  head(cars) # Display the first few records of a dataset
```

```{r}
  names(cars) # Display variable names of a data frame, one kind of data in R
```

```{r}
  cars$weight # Look at only the weight variable within data frame cars
```  

### Matrices
#### Commentary: The commands below creates a matrix based on the specified number of columns, rows, and values.
```{r}
  # Create a matrix with three rows, two columns, and every value equal to 0.0
  mat <- matrix(0.0, nrow = 3, ncol = 2); mat
  colnames(mat) <- c("Var 1", "Var 2") # Give a matrix variable names
  colnames(mat) # Display variable names of a matrix
```

### Subset data and declare new variables
#### Commentary: The commands below creates subsets within data that has been read into the system and creates new variables.
```{r]}
  cars.rsub <- cars[1:50,] # Subset the data by rows
  cars.csub <- cars[,1:3] # Subset by columns
  cars.rcsub <- cars[c(1,3,5), c(2,4)] # Subset by specific rows and columns
  cars.vsub <- cars[which(cars$mpg> 30),] # Subset by a logical condition
# To declare new variables, type the variable name, a left-arrow, then the value of the variable
  firstletter <- 'a'
  weight <- cars$weight
```

### Display more than one figure at a time
#### Commentary: The commands below plots figures based on the number of specified rows and figures within each row.
```{r}
  par(mfrow=c(1,1)) # plots one figure; the default setting
  par(mfrow=c(2,3)) # plots six figures: three in the top row, three in the bottom row
  # Plots will fill the plot space row by row
```  

### Download and install an R Package
#### Commentary: The command below installs/retrieves the ggplot2 package to enable the user to plot data in a graph.
```{r}
  #install.packages("ggplot2")
  library(ggplot2)
```  
  
# Chapter 2
  
### READ IN THE CARS AND CARS2 DATASETS
```{r}
  setwd("~/IMGT680") # sets the working directory
  cars <- read.csv("cars.txt", stringsAsFactors = FALSE) # reads text file Cars.txt
  cars2 <- read.csv("cars2.txt", stringsAsFactors = FALSE) # reads text file Cars.txt
```  

### MISSING DATA
```{r}
  # Look at four variables from cars
  cars.4var <- cars[, c(1, 3, 4, 8)] # renames data from columns 1,3,4,8 to cars.4var
  head(cars.4var) # displays first few records of each column
```
```{r}
  # Make certain entries missing
  cars.4var[2,2] <- cars.4var[4,4] <- NA # sets data from row 2, column 2 and row 4, column 4 to "NA".
  head(cars.4var)
```
```{r}  
  # Replace missing values with constants
  cars.4var[2,2] <- 0 # sets data from row 2, column 2 to zero
  cars.4var[4,4] <- "Missing" # sets data from row 4, column 4 to "Missing"
  head(cars.4var)
```
```{r}  
  # Replace values with mean and mode
  cars.4var[2,2] <- mean(na.omit(cars.4var$cubicinches)) # sets data from row 2, column 2 to the mean for CubicInches (201.5346)
  our_table <- table(cars.4var$brand) # sets data from "brand" column to new variable, our_table
  our_mode <- names(our_table)[our_table == max(our_table)] # sets the mode of our_table to new variable, our_mode
  cars.4var[4,4] <- our_mode # sets row 4, column 4 to the the mode (US)
  head(cars.4var)
```
#### Commentary: The commands below
```{r} 
  # Generate random observations
  obs_brand <- sample(na.omit(cars.4var$brand), 1) # generates 1 random number for brand columnand sets to new variable, obs_brand
  obs_cubicinches <- sample(na.omit(cars.4var$cubicinches), 1) # generates 1 random number for cubicinches column and sets to new variable obs_cubicinches
  cars.4var[2,2] <- obs_cubicinches # set row 2, column 2 to obs_cubicinches
  cars.4var[4,4] <- obs_brand # set row 4, column 4 to obs_brand
  head(cars.4var)
```

### CREATE A HISTOGRAM
```{r}
  # Set up the plot area
  par(mfrow = c(1,1)) # creates a multi-paneled plot with 1 row by 1 column
 
  # Create the histogram bars
  hist(cars2$weight, # extracts data from weight column in cars2.txt data
       breaks = 30, # sets bin width to 30
       xlim = c(0, 5000), # sets x-axis from 0 to 50000
       col = "blue", # outputs blue colored columns
       border = "black", # outputs a black border around each column
       ylim = c(0, 40), # sets y-axis from 0 to 40
       xlab = "Weight", # labels x-axis
       ylab = "Counts", # labels y-axis
       main = "Histogram of Car Weights") # creates graph title


  # Make a box around the plot
  box(which = "plot", lty = "solid", col = "black") # creates a black box border around the plot
```

### CREATE A SCATTERPLOT
```{r}
  plot(cars2$weight, cars2$mpg, # extracts data from weight and mpg columns in cars2.txt data
       xlim = c(0, 5000), ylim = c(0, 600), # sets x-axis from 0 to 5000
       xlab = "Weight", ylab = "MPG", # labels x-axis
       main = "Scatterplot of MPG by Weight", # creates graph title
       type = "p", pch = 16, col = "blue") # creates circular points filled in with the color blue
  #Add open black circles
  points(cars2$weight, cars2$mpg,
         type = "p", col = "black") # creates circular points filled in with the color black
```
 
### DESCRIPTIVE STATISTICS
```{r}
  mean(cars$weight) # Mean
  median(cars$weight) # Median
  length(cars$weight) # Number of observations
  sd(cars$weight) # Standard deviation
  summary(cars$weight) # Min, Q1, Median, Mean, Q3, Max
```
 
### TRANSFORMATIONS
```{r}
  # Min-max normalization
  summary(cars$weight)
  mi <- min(cars$weight) # minimum number
  ma <- max(cars$weight) # maximum number
  minmax.weight <- (cars$weight - mi)/(ma - mi) # min-max normalization
  minmax.weight
  
  # Z-score standarization
  m <- mean(cars$weight) # mean
  s <- sd(cars$weight) # standard deviation
  z.weight <- (cars$weight - m)/s # standard normal deviation
  z.weight
  length(cars$weight) # length of vector
  
  # Decimal scaling
  max(abs(cars$weight)) # 4 digits
  d.weight <- cars$weight/(10^4) # scaled by 10^4
  d.weight
```
  
### SIDE-BY-SIDE HISTOGRAMS
```{r}
  par(mfrow = c(1,2)) # creates a multi-paneled plot with 1 row by 2 columns
  
  # Create two histograms
  # Creates a histogram of weight
  hist(cars$weight, breaks = 20, xlim = c(1000, 5000),
       main = "Histogram of Weight", xlab = "Weight", ylab = "Counts")
  box(which = "plot", lty = "solid", col = "black")
  # Creates a histogram of weight using z-score
  hist(z.weight, breaks = 20, xlim = c(-2, 3),
       main = "Histogram of Z- score of Weight",
       xlab = "Z-score of Weight", ylab = "Counts")
  box(which = "plot", lty = "solid", col = "black")
```
  
### SKEWNESS
```{r}
  (3*(mean(cars$weight) - median(cars$weight)))/sd(cars$weight) # skewness formula for cars data
  (3*(mean(z.weight) - median(z.weight)))/sd(z.weight) # skewness formula for z-score of cars data
```
 
### TRANSFORMATIONS FOR NORMALITY
```{r}
  sqrt.weight <- sqrt(cars$weight) # Square root
  sqrt.weight_skew <- (3*(mean(sqrt.weight) - median(sqrt.weight))) / sd(sqrt.weight) # skewness formula for square root of weight data
  ln.weight <- log(cars$weight) # Natural log
  ln.weight_skew <- (3*(mean(ln.weight) - median(ln.weight))) / sd(ln.weight) # skewness formula for natural log of weight data
  invsqrt.weight <- 1 / sqrt(cars$weight) # Inverse square root
  invsqrt.weight_skew <- (3*(mean(invsqrt.weight) - median(invsqrt.weight))) /sd(invsqrt.weight) # skewness formula for natural log of weight data
```
 
### HISTOGRAM WITH NORMAL DISTRIBUTION OVERLAY
```{r}
  par(mfrow=c(1,1)) # creates a multi-paneled plot with 1 row by 1 column
  x <- rnorm(1000000, mean = mean (invsqrt.weight), sd = sd(invsqrt.weight)) # generates normal random numbers using mean and standard deviation
  hist(invsqrt.weight, 
       breaks = 30, 
       xlim = c(0.0125, 0.0275),
       col = "lightblue", prob = TRUE, border = "black",
       xlab = "Inverse Square Root of Weight",
       ylab = "Counts", main = "Histogram of  Inverse Square Root of Weight")
  box(which = "plot", lty = "solid", col="black")
  
  # Overlay with  Normal density
  lines(density(x), col = "red")
```
 
### NORMAL Q-Q PLOT
```{r}
  qqnorm(invsqrt.weight, datax = TRUE, col = "red", ylim = c(0.01, 0.03), # generates normal Quantile-Quantile (QQ) plot of the inverse sqrt values
         main = "Normal Q-Q Plot of Inverse Square Root of Weight")
  qqline(invsqrt.weight, col = "blue", datax = TRUE) # overlays theoretical QQ line
```
  
### DE-TRANSFORM DATA
```{r}
  # Transform x using y = 1 / sqrt(x)
  x <- cars$weight[1]; y <- 1 / sqrt(x)
  
  # Detransform x using x = 1 / (y)^2
  detransformedx <- 1 / y^2
  x; y; detransformedx
```
 
### CREATE INDICATOR VARIABLES
```{r}
  north_flag <- east_flag <- south_flag <- c(rep(NA, 10)) # combines replications of the the value "NA" 10 times
  region <- c(rep(c("north", "south", "east", "west"),2), "north", "south") # combines replications of N/S/E/W 2 times, North, and South
  
  # Change the region variable to indicators
  # Utilizing looping and if/else statements to set N/E/S to zeros and ones
  for (i in 1:length(region)) {
    if(region[i] == "north") north_flag[i] = 1
    else north_flag[i] = 0
    if(region[i] == "east") east_flag[i] = 1
    else east_flag[i] = 0
    if(region[i] == "south") south_flag[i] = 1
    else south_flag[i] = 0
  }
  north_flag; east_flag; south_flag
  
### INDEX FIELDS
  # Data frames have an index field;
  # the left-most column
  cars
  cars[order(cars$mpg),]
  
  # For vectors or matrices,
  # add a column to act as an index field
  x <- c(1,1,3:1,1:4,3); y <- c(9,9:1)
  z <- c(2,1:9)
  matrix <- t(rbind(x,y,z)); matrix # rbind = row bind
  indexed_m <- cbind(c(1:length(x)), matrix); indexed_m # cbind = column bind
  indexed_m[order(z),]
```  

### DUPLICATE RECORDS
```{r}
  # For number of duplicate records, use anyDuplicated
  anyDuplicated(cars)
  
  # To examine each record, use Duplicated
  duplicated(cars)
  # ‘True’: record is a duplicate,
  # ‘False’: record is not a duplicate
  
  # Let's duplicate the first record
  new.cars <- rbind(cars, cars[1,])
  
  # Check for duplicates
  anyDuplicated(new.cars)
  
  # The 262nd record is a duplicate
  duplicated(new.cars)
```
